Synthesizing Quotient Lenses

### 2018-03-08 papers/QuotientLenses/main.tex
Made changes

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
While there currently exist many bidirectional programming languages, many of
them do not have built-in support for transformations that operate on data
modulo equivalence relations such as whitespace or permutation
quotienting.\bcp{I'm worried that most readers will not be able to follow
  that.  Moreover, this way of introducing the paper feels kind of
  negative---most languages are insufficient, and Boomerang is hard to
  program.  Could we say it in a more accessible, more positive way?}
Moreover, langauges such as Boomerang which do allow for such transformations
are often difficult to program \sam{why?}. To address these shortcomings, this
paper (1) introduces a language of {\em quotient regular expressions} or QREs
for defining equivalences on regular expressions as well as a language of {\em
QRE lenses}, which are lenses that map between QREs, and (2) shows how to {\em
synthesize} QRE lenses from a pair of QREs and a set of example input-output
pairs. We also describe an implementation of QREs and QRE lens synthesis in
Boomerang, and demonstrate the utility of our implementation by using it to
synthesize QRE lenses between a variety of real-world data formats used by
the {\tt data.gov} database. 
\end{abstract}

@@

\section{Introduction}

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object.\bcp{These examples show that the same data {\em could} be
  represented in different formats, not that it needs to be.} When different applications manipulate the same data in different formats,
synchronizing the applications\bcp{Why are we talking about synchronizing
  applications?  Comes out of the blue.} amounts to synchronizing their
respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}\bcp{why mention exactly these two, out of all the
  bidirectional languages out there?}, typically come with a variety of
guarantees\bcp{grammatically, that says that each lens comes with a variety
  of guarantees---is that what you meant?},
the \emph{lens laws}, that help\bcp{why ``help''?} ensure put and get
functions are effective 
inverses\bcp{what is an effective inverse?} of one another.

One way\bcp{Do you mean to highlight the suggestion that there are other
  ways too?} to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
In bibliographies, for example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation\bcp{actually,
  two: one on each side; but more importantly ``introducing an equivalence
  relation'' is rather vague.} and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.

Ordinary bijective lenses are fairly tricky to program\bcp{do we have
  concrete evidence for this?}, especially for
beginners\bcp{or this?}, as they require programmers to think in two
directions at once\bcp{my own experience suggests that this is the less
  important issue...}
and learn a new type system, which comes with some tricky ambiguity
constraints\bcp{...and this is the bigger one.  But we don't explain what
  these ambiguity constraints are or why they must be imposed.}.
Unfortunately, quotient lenses do not improve the situation\bcp{Again, it
  feels rather negative.}:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions} (QREs), which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately---by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses}, and, we show how to synthesize such QRE
lenses automatically.
More specifically, in
order to generate a lens, a programmer provides a pair of QREs
describing the source and target formats, plus 
a (possibly empty) suite of examples that demonstrates how the
lens should act in specific cases.  Given this information,
we can exploit and extend recent work on algorithms for synthesis of
bijective lenses~\cite{optician} to generate quotient lenses.
The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

The following list summarizes the central contributions of our
work.\bcp{Ugh.} 
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\section{QRE Lenses by Example}
\label{sec:example}

%In this section, we explain the difficulties of programming with quotient
%lenses in Boomerang and show, by way of example, how our
%{\em Quotient Regular Expressions} (QREs) help alleviate these difficulties.

\emph{Lenses} define computations that can be read in two directions, the
get direction (forwards) and the put direction (backwards).  The simplest
lenses define bijections between source and data, but we are interested in
\emph{quotient lenses}, which define bijections modulo some equivalence
relation\bcp{two!}.  The core ideas of quotient lenses were developed by
Foster et al~\cite{quotientlenses} and are embodied in the Boomerang
language.  This section describes and motivates a 
new notation for defining such lenses.

\bcp{We need to write two new subsections here --- one on plain old
  bijective lenses, and one on bijective lens synthesis.}

\subsection{Quotient Lenses}

To begin, consider a bidirectional transformation that converts
\bibtex{} citation records such as
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway
\end{verbatim}
\noindent
and vice versa.
%
Notice that transformations from one format to the other preserve
most of the information present in the data:  the title, the author,
and the label, for instance.  However, some of the syntactic information
may be lost.  For instance, users are typically uninterested in preserving
whitespace between words or
between fields.  The order of author and title fields is also likely
irrelevant and there may be equivalent ways of writing the same
name:  ``Conway, J. H.'' vs ``J. H. Conway.''

%% Moreover,
%% if two \bibtex{} files differ only in ``inessential'' details, we would like
%% to treat them the same by mapping them to identical EndNote representations.
%% In this , the transformation should allow whitespace characters to occur
%% arbitrarily between the words in the label, author, and title fields, as well as
%% in between them. Second, an author's name can have the surname followed by a
%% comma and then the other names separated by arbitrary whitespace, or just all
%% the names separated by arbitrary whitespace (for example, the names ``Conway, J.
%% H.'' and ``J. H. Conway'' are equivalent). Lastly, the transformation should
%% also ignore the order in which the author and title fields occur in each
%% citation record.

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them 
into canonical representations.
Continuing our example, the following two
\bibtex{} citations represent the same logical object and as a result, should
be translated to the same EndNote record.
\begin{verbatim}
@Book {conway,
Author = "J. H. Conway",
Title = {Regular Algebra and Finite Machines},
}

@Book{     conway,
Title = {Regular Algebra
and Finite Machines},
Author = "Conway, J. H.",     }
\end{verbatim}

One natural way to define lenses that will synchronize data in these formats
is to provide the following items: \dpw{this is slightly redundant from the intro.  It
adds a little more info, but not much.  Should this be cut?  Should the intro be expanded
slightly?  Should it be left as is (more or less)?} \saz{I think that this is
not too redundant and that it can be left as-is, modulo space crunching.}

\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
    \bibtex{} and EndNote.
  Such descriptions may be given by regular expressions, for instance.
  They serve as documentation to describe the domains over which the lens
  is to operate.\bcp{This way of saying it implies that they serve {\em
      only} as documentation.}
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats, again, to document the structure of the
  transformation.\bcp{Same question.}
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.\bcp{Is it really true
  that these types are optional in Boomerang?}  In this case, there is 
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.\bcp{Would it be worth
  demonstrating this concretely?}  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and between,
fields while also ignoring field order, one must nest canonizing combinators
in a somewhat tricky way.

%% %Foster's Boomerang~\ref{boomerang} language presents one design
%% While it is possible to express the equivalence on the \bibtex{} data above in
%% Boomerang, actually doing so is a challenge.  Three of the inessential details
%% we wish to ignore---whitespace quotienting
%% inside fields, whitespace quotienting in between fields, and permutation
%% quotenting of fields---nest within each other in a way that makes them difficult
%% to express with the existing Boomerang combinators. Furthermore, if the
%% programmer wishes to specify the type of the lens, then she can only specify
%% data formats on each side of the lens with a regular expression that does not
%% keep track of the equivalence relation defined on the language of the regular
%% expression. Consequently, the programmer has to write a regular expression as
%% well as a canonizer that expresses the equivalence relation on that regular
%% expression, even though a regular expression and the canonizer defined on it
%% typically share much of the syntactic structure. Having to write both of these
%% expressions makes programming with quotient lenses verbose and time consuming.

We propose to do away with much of this overhead, and the burden of a
fiddly programming process, by defining a single new compact notation
to specify data formats modulo an equivalence relation and then by
synthesizing the lenses that map between these formats automatically.
\bcp{If our main contribution is ``eliminating fiddliness,'' we should be
  more explicit about what the fiddliness is.  (But, again, I would be even
  happier if we could phrase everything more positively... ``Quotient lenses
  make lens programming more realistic.  But here are some opportunities for
  eliminating redundancy...'')}

\subsection{Quotient Regular Expressions}
\label{subsec:qre-expressions}

A {\em Quotient Regular Expression} (QRE) defines an external regular
expression $R$, an equivalence relation $\sim$ on the language described by
$R$, and a canonical set of representatives of $\sim$, all at once.  QREs
are constructed from a simple set of primitives and composed using familiar
regular operators (`\cd{|}' for choice, `\cd{.}' for concatenation, and
`$^*$' for repetition).  Users\bcp{This sentence feels clunky because it
  ``uses up'' its head/subject position, and needlessly focuses the reader's
  attention, on something (``users'') that is not especially important or
  germane to the conversation.  (There are lots of other sentences elsewhere
  in the paper that do the same thing. :-) Please rewrite (all of them).}
may also treat QREs as pairs of get and put functions and compose such pairs
(the functional composition operator `;').\saz{I agree with Benjamin about
  useless subjects, but here the sentence could be deleted entirely.}

For example, to begin construction of our \bibtex{} format, we might
first define a whitespace format, which, externally, matches
any non-zero number of whitespace characters, and converts
any such whitespace into a single space character (its canonical form).
This whitespace-normalizing QRE is easily defined by using the \kw{project} primitive: \bcp{Why is it called ``project''?}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second. 
For instance, assume that
the \codefont{comma\_name} format describes ``Conway, J. H.''
and that the \codefont{space\_name} format describes ``J. H. Conway''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{s\_to\_c} canonization function used by squash
is simply to write it from scratch in some ordinary programming language. (Boomerang canonizers are written in OCaml.)
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{s\_to\_c} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
\bcp{Four problems here.  The first is that the notion of synthesizing lenses
comes out of nowhere (of course, it is in the background, but we're using it
here as if it had already been introduced concretely---this will be helped
by adding the new subsection I've suggested at the top of this section).
The second is that the idea of using half of a lens 
as a canonizer comes out of nowhere.  The third is that this idea is a hack (and
always was).  Why bother defining / synthesizing a lens and then using only
half of it?  And the fourth is that, in the original quotient lens paper,
we used this trick on lenses that were not even bijective, which may confuse
people familiar with the earlier paper.}The first line above synthesizes a
\emph{bidirectional}\bcp{Why is this italicized?  Are some lenses not
  bidirectional?} lens between 
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator ``;''.  For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \kw{squash} QRE and a QRE that projects away
the extra whitespace in between \codefont{first\_name} and \codefont{last\_name}

\begin{figure}[t]
\begin{lstlisting}
(* whitespace with " " as a canonical representative *)
let wsp = project [ \n\t\r]$^+$ $\to$ " "

(* define name representations with a space and with a comma *)
let space_name = name | (name . wsp . (name . wsp)$^*$ . name)
let comma_name = name | (name . "," . wsp . (name . wsp)$^*$ . name)

(* synthesize a lens that maps the comma representation to the space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l

(* squash QRE maps comma_name to space_name *)
let name = squash comma_name $\to$ space_name using s_to_c

(* define rest of bibtex fields *)
let name_sep = wsp . "and" . wsp
let bib_names = name | (name . name_sep . (name . name_sep)$^*$ . name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp . (word . wsp)$^*$ . word)
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}

\caption{QRE specification of \bibtex{} records. \bcp{The spacing of the
    comments is weird...}}
\label{fig:example-qre}
\end{figure}

@@

% The final QRE combinator is the \kw{normalize} combinator.  This combinator is an
% ``escape hatch'' that allows a programmer to manually define a function $f$
% that sends each string matching a regular expression $R$ to some canonical
% representative in another regular expression $R'$. The equivalence relation
% defined by the normalize combinator is hence the equivalence relation defined by
% the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that match $R$,
% $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

>>>>

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
There currently exist many bidirectional programming languages, some of which
allow a programmer to define a transformation that respects an equivalence
relation. For example, a language with this feature may allow a programmer to
define a transformation which ignores the order of fields in an XML file, so
that two files that have the same fields with the same data but with the fields
occuring in a different order are considered the same by the
transformation. This paper simplifies the task of programming such {\em
quotient lenses} in two main ways. First, we introduce {\em
quotient regular expressions} or QREs which are regular expressions
augmented with extra syntax that makes it easier to specify equivalence
relations on regular langauges than currently existing tools. Secondly, we
introduce {\em QRE lenses}, which are lenses that map between QREs, and show
how to {\em synthesize} QRE lenses from a pair of QREs and a set of example
input-output pairs. We also describe an implementation of QREs and QRE lens
synthesis in the bidirectional programming language Boomerang, and demonstrate
the utility of our implementation by using it to synthesize QRE lenses between
various real-world data formats used by the {\tt data.gov} database.
\end{abstract}

@@

\section{Introduction}
Oftentimes, the same data is stored in a variety different formats. For
example, bibliography data may be stored in \bibtex{} and
EndNote formats; spreadsheet data can be represented as a CSV file
or a TSV file; web APIs can return data as a JSON or XML object. In such cases,
it is necessary to synchronize the two or more copies of data so that they
store the same information.

One way to synchronize data stored in different formats is to use \emph{lenses},
\emph{i.e.,} programs that represent pairs of functions for translating data
from format A to B (the \emph{get} direction) and back again (the \emph{put}
direction). Lenses are usually derived using domain-specific
bidirectional languages which guarantee that the lenses defined in that language
satisfy a certain set of \emph{lens laws} that give guarantees on how the two
transformations interact.

Amongst the most restrictive lenses are the \emph{bijective lenses}.  These
lenses demand that get and put functions form a bijection, meaning no
information is lost when one translates from one data format to another.  While
this strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant. Many
data formats have other ad hoc rules as well. In bibliographies, for example,
recall that in \bibtex{}, an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not
matter which of the two formats is used.

{\em Quotient lenses} loosen many of the restrictions of bijective lenses by
allowing a programmer to define an equivalence relation on the source and
target formats so that the round-tripping lens laws hold modulo these
equivalence rather than modulo equality. Quotient lenses thus make it possible
to align many more data sources than would otherwise be possible, by ignoring
details that a programmer views as inessential. The only language we are aware
of that implements a general class of quotient lenses is {\em
Boomerang}~\cite{quotientlenses}. While Boomerang's quotient lenses enable a
programmer to define a very general class of quotient lenses, these lenses to
have two main drawbacks.

First, in order to define an equivalence relation in Boomerang, a programmer
must define a {\em canonizer}; pair of functions $f$ and $g$ where $f$ maps
data represented in the original external format to a canonical element and $g$
chooses which elements of the original format correspond to values in the
canonical format. The disadvantage of this approach is that in order for a
programmer to {\em specify} a quotient lens in Boomerang, she must define four
different data formats: two ``outer'' formats which specify the data that is to
be transformed by the lens, and two ``inner'' formats which specify the
canonical formats for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} or QREs, which enable a programmer to simultaneously specify a
regular expression as well as a set of canonical representatives for strings
that match the regular expression. QREs have the added advantage that once the
programmer has defined a well formed QRE, then she need not define a
canonization function seperately; the canonization function can automatically be
inferred from the QRE.

The second challenge with programming lenses in Boomerang is that even after the
programmer has specified the data as well as a canonizer that maps each member
in the source and target sets to their respective canonical representative, she
must still define the underlying lens that maps between the respective sets of
canonical representatives. This can be a slow and tedious task due to the
constraints that Boomerang imposes on its terms. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

Our solution to this problem is that we use QREs to define a class of quotient
lenses called {\em QRE lenses}, and, we show how to synthesize such QRE lenses
automatically. More specifically, in order to generate a lens, a programmer
provides a pair of QREs describing the source and target formats, plus
a (possibly empty) suite of examples that demonstrates how the
lens should act in specific cases.  Given this information, we can exploit and
extend recent work on algorithms for synthesis of bijective
lenses~\cite{optician} to generate quotient lenses. The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for simultaneously defining a
  regular language modulo some equivalence relation and a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that each QRE lens is functionally equivalent to a QRE lens in the
  canonical form of a bijective lens with ``canonizers at the edges''.
   (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Doe
%A John Doe
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

{\em Bijective lenses} are designed to define a bidirectional transformation
such as this, where the data formats can be matched in a one-to-one manner, in
this case by matching the label, author and title fields. Boomerang allows
programmers to define such bijecive transformations using combinators which
favour a compositional approach in which the programmer defines small lenses
which are then put together using the combinators to define larger and larger
lenses until the final lens is defined.

Each bijective lens is constructed from simple primitive combinators including:
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
del ",\nauthor = \""
. ins "\n%A "
. name
. (del " and "
. ins "\n"
. (ins "%A " . name . del " and " . ins "\n")*
. ins "%A "
. name
|| "")

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between \cd{bibtex} and \cd{end_note} regular
expressions with no equivalences.\bcp{And a readability suggestion:
define a named constant for a quote character.}}
\label{fig:example-lens}
\end{figure}

Unfortunately, programming bijective lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behaviour of the lens on a few inputs in order to define a
bijective lens. As a result defining bijective lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective lens rather
than manually derive the lens herself.

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Doe, John'' vs ``John Doe.''

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them into canonical representations.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title},
}

@Book{     Doe,
Title = {Generic
Title},
Author = "Doe, John",     }
\end{verbatim}

One natural way to define lenses that will synchronize data in these formats
is to provide the following items:
\begin{enumerate}
  \item A pair of descriptions of the external source and target formats, here
  \bibtex{} and EndNote. Such descriptions may be given by regular
  expressions, for instance.
  \item A pair of \emph{canonizers} that map external source (and target)
  objects into internal, canonical forms.  Such canonizers will take
  care of stripping away irrelevant information and otherwise normalizing
  differences between equivalent representations of the same logical data.
  When given an internal representation, such canonizers must also operate
  in the reverse direction, \emph{choosing} an element in the external format
  to represent the canonical internal element.
  \item A pair of descriptions of the internal source and
  target formats.
  \item A bijective lens to map back and forth between canonical elements
  of the internal source and target formats.
\end{enumerate}

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}. Of course, some of these elements are optional.  For
instance, it is not necessary to document internal and external formats.
However, we do view it as good engineering practice to do so, as such
documentation makes lenses easier to read, understand, use and maintain.

The first way in which our work improves upon Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe the QRE combinators and use give an example of how a
programmer might define a QRE specification of \bibtex.

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
To construct a QRE specification of a \bibtex{} format, we might first define a
whitespace format, which, externally, matches any non-zero number of whitespace
characters, and converts any such whitespace into a single space character (its
canonical form). This whitespace-normalizing QRE is easily defined by using the
\kw{project} primitive: \bcp{Why is it called ``project''?}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Doe, John''
and that the \codefont{space\_name} format describes ``John Doe''
and \codefont{s\_to\_c} is a function from the first to the second.  In this case,
the following instance of \kw{squash} creates the desired canonizer.

@@

One way to define the \codefont{s\_to\_c} canonization function used by squash
is simply to write it from scratch in some ordinary programming language. (Boomerang canonizers are written in OCaml.)
However, our system will also synthesize such functions automatically from pairs
of regular languages and examples---here, \codefont{s\_to\_c} is the
\codefont{get} direction of a lens that can be synthesized by:
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l
\end{lstlisting}
%
\bcp{Three problems here. First, the idea of using half of a lens
as a canonizer comes out of nowhere.  The second is that this idea is a hack
(and always was).  Why bother defining / synthesizing a lens and then using only
half of it?  And the third is that, in the original quotient lens paper,
we used this trick on lenses that were not even bijective, which may confuse
people familiar with the earlier paper.}

\sam{Here is my response to this. You observe that the idea of using half a lens
as a canonizer comes out of nowhere, and that this ideas is a hack and always
was. I don't think that's quite right. I think that it is useful to have a
combinator which chooses between two or more formats. Indeed, the permutation
combinator is just a special case of this. The bibtex name field is a good
example of where such a combinator would be useful.  But if we choose to have
such a combinator, then we need a way of mapping one format to another, and
really the only interesting string-to-string transformations in Boomerang are
given by lenses, which is why we use the get function of a lens in a canonizer.
We could have used any other function between the two formats but Boomerang
doesn't give us too many options for writing unidirectional transformations. So
I don't think that using a lens as a canonizer is a trick. I think that
Boomerang just isn't designed for unidirectional tranformations. But I also
agree that it doesn't make much sense to synthesize a lens only to get rid of
one half of it.
I'm not sure what the solution is to be honest.}

The first line above synthesizes a
lens between \codefont{comma\_name} and \codefont{space\_name} using the listed
example transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for \kw{squash}.

Another QRE primitive is the sequential composition combinator ``;''.  For
example, suppose that we wish to express a name as $\kw{last\_name \cdot
``,\text{''} \cdot \text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot
\text{wsp+} \cdot last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$ as the canonical representation. We also wish to project away the
extra whitespace in between \kw{first\_name} and \kw{last\_name} once this
identification has been made. Then we can express this compound equivalence
relation by a composition of a \kw{squash} QRE and a QRE that projects away
the extra whitespace in between \codefont{first\_name} and
\codefont{last\_name}.

@@

The final QRE combinator is the normalize combinator.
This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p\_wsp} which matches any non-zero number of whitespace
characters and converts any such whitespace into a single space character can
be expressed using the normalize combinator as 
\lstinline{normalize (wsp+, " ", f)}.

Figure~\ref{fig:example-qre} gives the complete QRE specification for the
\bibtex{}.
\begin{figure}
\begin{lstlisting}
let wsp = [ \n\t\r]

(* define name representations with a space and with a comma *)
let space_name = first_name . wsp+ . last_name
let comma_name = last_name . "," . wsp+ . first_name

(* synthesize a lens that maps comma representation to space representation *)
let l = synth comma_name $\Leftrightarrow$ space_name using {("Doe, John", "John Doe")}
let s_to_c = get l

(* squash QRE maps comma_name to space_name *)
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* map arbitrary whitespace into " " as a canonical representative *)
let p_wsp = project wsp+ $\to$ " "

(* Project away whitespace in space_name representation *)
let project_in_space = first_name . p_wsp . last_name
